\chapter{Background material}
\label{chap:background-material}

In this chapter we overview a selection of concepts which we need
later on. We also fix notation and a number of definitions. At the
momement the sections are not listed in any particular order.

\paragraph{Free and bound variables.}

Occurrences of variables in an expression may be \emph{free} or
\emph{bound}. Variables are bound when they are used to indicate the
range over which an operator acts. For example, in expressions
%
\begin{equation*}
  \xall{x}{\RR}{x^2 + y \geq 0},
  \qquad\qquad
  \sum_{k = 0}^n \frac{1}{k^2},
  \qquad\qquad
  \int_a^b f(t) \, dt,
\end{equation*}
%
the variables $x$, $k$, and $t$ are bound by the operators $\forall$,
$\sum$, and $\int$, respectively. The remaining variables are free. It
is really the \emph{occurrence} of a variable that is bound or free,
not the variable itself. In
%
\begin{equation*}
  P(x) \lor \xusome{x}{\lnot Q(x)}
\end{equation*}
%
the left-most occurence of $x$ is free whereas the other two are bound
by $\exists$.

\paragraph{Functions.}

The set of all functions from $A$ to $B$ is denoted by $B^A$ or $A \to
B$. The arrow associates to the right, $A \to B \to C$ is $A \to (B
\to C)$. We write $f : A \to B$ instead of $f \in A \to B$. If $f : A
\to B$ and $x \in A$, the application $f(x)$ is also written as $f\,
x$. We often work with \emph{curried} functions which take several
arguments in succession, i.e., if $f : A \to B \to C$ then $f$ takes
$x \in A$, and $y \in B$ to produce an element $f(x)(y)$ in $C$, also
written $f\, x\, y$.


\paragraph{Partial functions.}

A \emph{partial} function\footnote{In the literature on Type Two
  Effectivity the common notation is $f \mathbin{{:}{\subseteq}} A \to
  B$.} $f: A \parto B$ is a function that is defined on a subset
$\dom{f} \subseteq A$, called the \emph{domain} of~$f$. Sometimes
there is confusion between the domain~$\dom{f}$ and the set~$A$, which
is also called the domain. In such cases we call $\dom{f}$ the
\emph{support} of~$f$. If $f: A \parto B$ is a partial function and $x
\in A$, we write $\defined{f\, x}$ to indicate that $f x$ is defined.
For an expression~$e$, we also write $\defined{e}$ to indicate
that~$e$ and all of its subexpressions are defined. The
symbol~$\downarrow$ is sometimes inserted into larger expressions, for
example, $\defined{f\, x} = y$ means that $f x$ is defined and is
equal to~$y$. If $e_1$ and $e_2$ are two expressions whose values are
possibly undefined, we write $e_1 \kleq e_2$ to indicate that either
$e_1$ and $e_2$ are both undefined, or they are both defined and
equal. The notation $e_1 \klgeq e_2$ means that if $e_1$ is defined
then $e_2$ is defined and they are equal. Thus we have
%
\begin{equation*}
  e_1 \kleq e_2 \iff e_1 \klgeq e_2 \land e_2 \klgeq e_1.
\end{equation*}

A partial map $f: X \parto Y$ between topological spaces~$X$ and~$Y$
is said to be \emph{continuous} when it is continuous as a total map
$f: \dom{f} \to Y$, where the domain of definition $\dom{f} \subseteq
X$ is equipped with the subspace topology. Note that there is no
restriction on the domain~$\dom{f}$.




\paragraph{Order theory.}

A \emph{preorder} $(P, {\leq})$ is a set with a reflexive and
transitive relation.

A \emph{partially ordered set (poset)} $(P, {\leq})$ is a set with a
reflexive, transitive, and anti-symmetric relation.

A function $f : P \to Q$ between posets is \emph{monotone} if $x \leq
y$ in $P$ implies $f(x) \leq f(y)$ in $Q$.

A subset $S \subseteq P$ is an \emph{upper set} if $x \in S$ and $x
\leq y$ implies $y \in S$. Similarly, it is a \emph{lower set} if $y
\in S$ and $x \leq y$ implies $x \in S$.

A subset $S \subseteq P$ of a poset $(P, {\leq})$ is \emph{directed}
if it is non-empty and for every $x, y \in S$ there exists $z \in S$
such that $x \leq z$ and $y \leq z$.

An \emph{upper bound} of a subset $S \subseteq P$ in a poset is an
element $x \in P$ such that $y \leq x$ for all $y \in S$.

The \emph{supremum} $\sup S$ of a subset $S \subseteq P$ in a poset is
its least upper bound, if it exists. More precisely, it is an upper
bound $x$ for $S$ such that if $y$ is also an upper bound for~$S$ then
$x \leq y$.

A \emph{directed-complete partial order (dcpo)} is a poset in which
every directed set has a supremum. Let $(D, {\leq})$ be a dcpo. For
$x, y \in D$ we say that~$x$ is \emph{way below} $y$, written $x \wayb
y$, when for every directed $S \subseteq D$ such that $y \leq \sup S$
there exists $z \in S$ for which $x \leq z$. An element $x \in D$ is
\emph{compact} (or \emph{finite}) when $x \wayb x$. A subset $U
\subseteq D$ is \emph{Scott open} if it is an upper set and is
inaccessible by suprema of directed sets, which means that, for every
directed $S \subseteq D$, if $\sup S \in U$ then already $x \in U$ fo
rsome $x \in S$. The Scott open sets form the \emph{Scott topology}
of~$D$.

If $D$ and $E$ are dcpos then a function $f : D \to E$ is continuous
with respect to the Scott topologies precisely when it preserved
suprema of directed sets. It follows that such a function is monotone.




%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "notes"
%%% End: 
