\documentclass[a4paper,11pt]{article}

\usepackage{a4wide}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{theorem}

\input{../macros}

\begin{document}

\title{Izračunljivost v topologiji}
\author{Homework 2}
\date{Due date: April 9th, 2009}

\maketitle


\subsection*{Instructions}

You are \emph{encouraged} to use the literature and research papers,
but any mathematical text that your solution relies on must come from
a peer-reviewed source (published book or journal article).

If you use information from Wikipedia or random PDF files on the
internet, then the correctness of the source becomes \emph{your}
responsibility, i.e., you should personally verify the proofs or write
them if they are missing.

Thus you may solve a problem, or part of a problem, by finding the
solution in a book or journal article. In this case you may simply
refer to the text you found (there is no need to copy it). If a
problem asks for a proof, you must refer to an actual proof (not a
paper which claims that the proof is trivial or left as exercise).

I prefer a hand-written solution to one written with Microsoft Word in
which mathematics is messed up. You may but do not have to use
{\LaTeX}.

If the problems are too hard, come to me and we will discuss them.

\subsection*{Problem 1}

Suppose $\AA$ is a TPCA and $\compAA$ a sub-TPCA. Let $t$ be a type
and $r_0 \in \compAtyp{t}$. Define the functor $\nabla' : \Set \to
\AsmA$ as follows. Given a set $X$, let $\nabla' X = (X, \Atyp{t},
{\rz_{\nabla' X}})$ with
%
\begin{equation*}
  r \rz_{\nabla' X} x \iff r = r_0.
\end{equation*}
%
A function $f : X \to Y$ is mapped to $\nabla' f = f$, which is
obviously realized.

Recall that we defined $\nabla X = (X, \Atyp{t}, {\rz_{\nabla X}})$
where $r \rz_{\nabla X} x$ for all $r \in \Atyp{t}$ and $x \in X$.

\begin{enumerate}
\item State the definition of ``functors $F$ and $G$ are naturally
  isomorphic''.
\item Verify that $\nabla$ and $\nabla'$ are naturally isomorphic.
\item Now suppose we picked $r_0 \in \Atyp{t}$ so that $r_0 \not\in
  \compAtyp{t}$. What would go wrong?
\end{enumerate}

\begin{center}
  (There are two more problems on page 2.)
\end{center}

\newpage

\subsection*{Problem 2}

A \emph{$0$-equilogical space} is an equilogical space $(X, {\equiv})$
such that $X$ is a $0$-dimensional countably-based space. This means
that we are given together with~$X$ a countable subbasis $(U_i)_{i \in
  \NN}$ of~$X$ such that every $U_i$ is both an open and a closed set.
Let $\mathsf{0Equ}$ be the category of $0$-equilogical spaces and
extensional maps between them.

Show that $\mathsf{0Equ}$ is equivalent to $\Asm{\Baire}$. Hint: show
that every $0$-dimensional countably-based space~$X$ may be
pre-embedded in $\Baire$, the embedding is similar to the
neighborhood-filter emebedding from the Embedding Theorem
for~$\Scott$. Then mimic the proof that $\Equ$ is equivalent to
$\Asm{\Scott}$.


\subsection*{Problem 3}

\begin{enumerate}
\item Prove that in any category with binary products $A \times B
  \cong B \times A$.
\item 
  Prove that in $\AsmA$ the following distributive law holds:
  % 
  \begin{equation*}
    (\asm{S} + \asm{T}) \times \asm{U} \cong
    (\asm{S} \times \asm{U}) + (\asm{S} \times \asm{U}).
  \end{equation*}
\item Find a category which has binary products and coproducts, but
  the distributive law does not hold.
\end{enumerate}

\end{document}
